Transition path sampling based calculations of free energies of enzymatic reactions: the case of human MAT2A and plasmodium vivax adenosine deaminase

### 2022-04-14 tex/manuscript.tex
added some more text

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jpcbfk,manuscript=article,layout=traditional]{achemso}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]
  {Transition path sampling based calculation of free energies of enzymatic
  reactions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{TPS,NMR,UV}
\keywords{American Chemical Society, \LaTeX}

@@

Calculating equilibrium properties such as the free energy of an enzymatic 
catalysis reaction within TPS is not straighforward. 
The Monte Carlo algorithm used in TPS has an acceptance criterion that 
is only satisfied if the trajectory begins in the reactant basin and ends in 
the product basin or vice versa, hence trajectories that are localized in the 
reactant or product basins are rejected. The localized trajectories contribute
to the probability density distribution function ($P(\lambda)$) which is the 
probability to find the system at a particular value of the order parameter $\lambda$.

Since the TPS path ensemble only consists of reactive trajectories, the information from 
the discarded non-reactive trajectories which is required for obtaining the 
equilibrium probability distribution of the order parameters is 
missing. \cite{Dellago09AdvCompSimAppp167} Radhakrishnan et al. developed 
`BOLAS' which combines the shooting and shifting algorithms of 
TPS with a window based sampling technique 
in the spirit of umbrella sampling to calculate equilibrium free energy 
differences. \cite{Radhakrishnan04JChemPhys121p2436} 
Baron peters developed a variant of this scheme called the 
equilibrium path sampling (EPS) where they use aimless shooting 
to create the ensemble within each window. \cite{Peters08JAmChemSoc130p17342,Silveira21JPhysChemB125p2018} 
Apart from these schemes a recent work by 
Brotzakis et al., \cite{Brotzakis19JChemPhys151p174111,Dellago04MolSimul30p795}
has attempted to look at the TPS based free energy calculations for 
reactions that are diffusive in nature where 
Transition interface sampling (TIS) \cite{vanErp03JChemPhys118p7762}
could work well.

In this manuscript we will focus on the implementation and application of the 
free energy based calculations within the TPS method on reactions catalyzed by two systems:
the human MAT2A enzyme and the plasmodium vivax adenosine deaminase 
enzyme. \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618}
These enzymes were chosen not only because of their importance in biochemical pathways and
but also because of the availability of experimental results with which comparisons can 
be drawn. 
%------------------------------------------------------------------------------ 
\section{Methods}
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------ 
Here we briefly describe the equations describing the TPS method for the purpose of
introducing notations as well as for the sake of being complete. 
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and 
momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is 
run for a total 
time of $\mathcal{T}$ the number of time slices is given by 
$L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

(7L1A from the protein data bank) reported by Niland 

@@

in complex with MT-coformycin (3EWC from the protein data bank) reported by Ho et 

@@

simulations. For the human MAT2A enzyme the substrates were modified to 
S-Adenosyl methionine (SAM) and adenosyl triphosphate (ATP) whereas for the \textit{pv}ADA

@@

MD simulations within classical and hybrid 
quantum mechanics/molecular mechanics (QMMM) approximations were carried out 
using the CHARMM program package. \cite{Brooks83JComputChem4p187,Brooks09JComputChem30p1545} 
For the MM calculations the CHARMM36 force field \cite{Brooks09JComputChem30p1545} 
was used and for the QM calculations the approximate PM3 semiempirical 
method \cite{Repasky02JComputChem23p1601} was used. Both the
systems were solvated with water molecules using the TIP3P model
in a nanodroplet, whose volume was set to be $15\;\AA$ away from
the surface of the protein, and then neutralized with potassium
ions.

Each protein was solvated within
a 15 $\AA$ sphere using TIP3 waters and neutralized using
potassium ions. We minimized the system with 50 steps of the
Steepest Descent method, followed by 2000 steps of the
adopted Basis Newton-Raphson method. The system was
heated slowly to 300 K for 30 ps, starting with harmonic
constraints except on H atoms and TIP3 waters that were
gradually reduced to force constant 1. We started the
equilibration with 15 ps with harmonic constraints of force
constant 1, which were then removed and the equilibration ran
for 60 ps.

We used transition path sampling to generate and
analyze the ensemble of reactive trajectories. As order
parameters that define the reactant and product basins, we
used the donor-hydride and acceptor-hydride distances. The
reactant basin was defined as a donor-hydride distance smaller
than 1.6  and the product basin as an acceptor-hydride
distance smaller than 1.6 . To seed the generation of the path
ensemble, we generated a biased trajectory by applying a
harmonic force between the hydride and the acceptor and
allowed the system to propagate for 250 fs. New trajectories
were generated by the shooting algorithm: we perturbed the
momenta of a randomly chosen slice of a reactive trajectory,
and using the new momenta, we allowed the system to
propagate forward and backward in time for 250 fs. A total of
120 reactive trajectories were generated.
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
%------------------------------------------------------------------------------ 
Human methionine adenosyltransferase (MAT2A) is an enzyme that catalyzes the 
reaction between adenosine triphosphaste and L-methionine which leads to the 
formation of S-Adenosylmethionine. \cite{Firestone17JAmChemSoc139p13754,Niland21Biochem60p791} 
The mechanism of this reaction is experimentally determined to follow a 
SN$_2$ transition state with the sulphur atom of the L-methionine as the 
nucleophile with the triphosphate leaving group.

@@

%------------------------------------------------------------------------------ 
\subsection{Adenosine deaminase reaction coordinates and free energy}
%------------------------------------------------------------------------------ 
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.22]{./figures/ada-protein.png}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[scale=0.6]{./figures/ada-fenergy.pdf}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jpcbfk,manuscript=article,layout=twocolumn]{achemso}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]
  {Transition path sampling based calculations of free energies of enzymatic
  reactions: the case of human MAT2A and plasmodium vivax adenosine deaminase}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{TPS,NMR,UV}
\keywords{American Chemical Society, \LaTeX}

@@

Calculating equilibrium properties such as the free energy of an enzymatic 
catalysis reaction within TPS is not straighforward. 
The Monte Carlo algorithm used in TPS has an acceptance criterion that 
is only satisfied if the trajectory begins in the reactant basin and ends in 
the product basin or vice versa, hence trajectories that are localized in the 
reactant or product basins are rejected. The localized trajectories contribute
to the probability density distribution function ($P(\lambda)$) which is the 
probability of a system at equilibrium to have a particular value of the 
order parameter $\lambda$. \cite{Dellago09AdvCompSimAppp167}

Frenkel has shown that inclusion of rejected states in Markov state Monte 
Carlo algorith can help speed up the convergence of the calculations of 
order-parameter distribution of a two-dimensional
Ising model. \cite{Frenkel04ProcNatAcadSci101p17571}
Inclusion of rejected trajectories within TPS was first discussed by Radhakrishnan 
et al. \cite{Radhakrishnan04JChemPhys121p2436} who developed `BOLAS' which combines 
the shooting and shifting algorithms of TPS with a window based sampling technique 
in the spirit of umbrella sampling to calculate equilibrium free energy 
differences. Peters et al. used a variant of the BOLAS scheme called the 
equilibrium path sampling (EPS) where they use aimless shooting 
combined with window based sampling for free energy calculations. \cite{Peters08JAmChemSoc130p17342,Beckham10epsbook} 
More recently Brotzakis et al., \cite{Brotzakis19JChemPhys151p174111}
have developed a reweighted path ensemble scheme within the standard TPS method 
called virtual interface exchange TPS (VIE-TPS) to calculate the free energy 
landscape. VIE-TPS is shown to produce reasonable free energies for 
diffusive reactions whereas it is less accurate otherwise.

Free energy calculations based on the TPS shooting algorithm is 
appealing but not a lot of examples exist in the literature 
which compare the computational calculations with experimental
results for enzymatic reactions.  
In this manuscript we implement and apply the 
algorithm developed by Radhakrishnan et al. \cite{Radhakrishnan04JChemPhys121p2436} 
to calculate the free energies of reactions catalyzed by two systems:
the human MAT2A enzyme and the plasmodium vivax adenosine deaminase 
enzyme. \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618}
These enzymes were chosen not only because of their importance in various 
biochemical applications but also because of the availability of experimental 
results with which comparisons can be drawn. The manuscript is organized as 
follows: first we discuss the TPS method and the algorithm that we use 
to calculate the free energies briefly, then we show calculations which apply this 
method on the two enzymatic reactions. Finally we compare our results to expeiments
and provide conclusions. 
%------------------------------------------------------------------------------ 
\section{Methods}
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------ 
Here we briefly describe the equations describing the TPS method for the purpose of
introducing notations as well as for the sake of being complete. 
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and 
momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is 
run for a total 
time of $\mathcal{T}$ the number of time slices is given by 
$L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

(protein data bank ID: 7L1A) reported by Niland 

@@

in complex with MT-coformycin (protein data bank ID: 3EWC) reported by Ho et 

@@

simulations. For the MAT2A enzyme the substrates were modified to 
L-methionine and adenosyl triphosphate (ATP) whereas for the \textit{pv}ADA

@@

MD simulations within classical and hybrid quantum mechanics/molecular mechanics 
(QMMM) approximations were carried out using the CHARMM program 
package. \cite{Brooks83JComputChem4p187,Brooks09JComputChem30p1545} 
For the MM calculations the CHARMM36 force field \cite{Brooks09JComputChem30p1545} 
was used and for the QM calculations the approximate PM3 semiempirical 
method \cite{Repasky02JComputChem23p1601} was used. 
We solvated the two proteins using TIP3P water molecules 
in a nanodroplet sphere whose volume was fixed at $15$ {\AA}
away from the protein's surface and the total charge was neutralized 
using potassium ions. Both the systems were minimized with 50 steps of the
steepest descent method, followed by 2000 steps of the
adopted basis Newton-Raphson method. The minimized systems were
heated slowly to 300 K for 35 ps beginning with harmonic
constraints on all atoms except on the H atoms and the TIP3 
waters with gradual reduction of the restraint forces to 1. 
15 ps of equilibration was carried out starting with harmonic 
constraints of force constant 1 followed by 20 ps of contraint free 
equilibration to prepare the systems for production runs.  
\subsection{Transition path sampling}
We used TPS to generate an ensemble of reactive trajectories. 
The time step that we used for all our calculations was $1\;fs$ with a total
of 250 steps forward and 250 steps backward which amounts to a total of $0.5\;ps$
trajectory.  
\subsubsection{human MAT2A}
As order parameters that define the reactant and product basins, we
used the donor-hydride and acceptor-hydride distances. The
reactant basin was defined as a donor-hydride distance smaller
than 1.6  and the product basin as an acceptor-hydride
distance smaller than 1.6 . To seed the generation of the path
ensemble, we generated a biased trajectory by applying a
harmonic force between the hydride and the acceptor and
allowed the system to propagate for 250 fs. New trajectories
were generated by the shooting algorithm: we perturbed the
momenta of a randomly chosen slice of a reactive trajectory,
and using the new momenta, we allowed the system to
propagate forward and backward in time for 250 fs. A total of
120 reactive trajectories were generated.

\subsubsection{plasmodium vivax ADA}
%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
For the free energy calculations we divided the reaction coordinate into 13 windows for 
MAT2A and 20 windows for pvADA with a $15\%$ overlap on each side of the window.

%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
%------------------------------------------------------------------------------ 
Human methionine adenosyltransferase (MAT2A) is an enzyme that catalyzes the 
reaction between adenosine triphosphaste and L-methionine which leads to the 
formation of S-Adenosylmethionine. \cite{Firestone17JAmChemSoc139p13754,Niland21Biochem60p791} 
The mechanism of this reaction is experimentally determined to follow a 
SN$_2$ transition state with the sulphur atom of the L-methionine as the 
nucleophile with the triphosphate leaving group.

@@

%------------------------------------------------------------------------------ 
\subsection{Adenosine deaminase reaction coordinates and free energy}
%------------------------------------------------------------------------------ 
\begin{figure}[ht!]
\includegraphics[scale=0.15]{./figures/ada-protein.png}
\end{figure}

\begin{figure}[ht!]
\includegraphics[scale=0.5]{./figures/ada-fenergy.pdf}
\end{figure}
